---
title: "Introducing Quantile Regression"
subtitle: "From Conditional Quantiles to the Regression Model"
author: "Tatjana Kecojević"
date: "24 June 2026"

bibliography: references.bib

format:
  revealjs:
    theme: serif
    slide-number: true
    transition: fade
    background-transition: fade
    footer: "Beyond the Mean · Session 1 · Tatjana Kecojević"
    logo: "UoM_logos/UoM_logo_TAB_col_white_background.png"
    smaller: false
    scrollable: false
    code-fold: true
    code-summary: "Show code"
    highlight-style: github
    fig-align: center

execute:
  echo: true
  warning: false
  message: false
  freeze: false
---

```{r setup}
#| include: false

library(tidyverse)
library(quantreg)
library(AER)

theme_set(theme_minimal(base_size = 14))

set.seed(4821)

check_fn <- function(u, tau) u * (tau - as.numeric(u < 0))
taus <- c(0.10, 0.25, 0.50, 0.75, 0.90)
```

# Part 3 {background-color="#2C3E50"}

## Learning Objectives

By the end of this session you will be able to:

::: incremental
- Explain what quantile regression estimates and how it extends OLS
- Describe the check function and the role of the indicator $\mathbf{1}(u < 0)$
- Read and run `rq()` in R
- Interpret a quantile regression coefficient at a specific $\tau$
- Explain when QR and OLS diverge, and why it matters
:::

# The Key Idea {background-color="#2C3E50"}

## OLS estimates one thing

OLS estimates the **conditional mean**:

$$E(Y \mid X) = X'\beta$$

One set of coefficients. One line through the data.

. . .

> *"A single mean curve might not be informative enough."*

## Quantile regression estimates the whole distribution

Quantile regression replaces the conditional mean with the **conditional quantile**: $Q_\tau(Y \mid X) = X'\beta(\tau)$

. . .

Two things change:

| | OLS | Quantile Regression |
|---|---|---|
| Estimates | Conditional **mean** | Conditional **quantile** |
| Coefficients | One set | One set **per** $\tau$ |
| Lines | One | A family |

## When does it matter?

::: {.columns}
::: {.column width="50%"}
**Parallel lines** $\rightarrow$ OLS tells the full story

All quantiles shift equally with $X$. The effect is *homogeneous*.
:::
::: {.column width="50%"}
**Fanning lines** $\rightarrow$ OLS misses the heterogeneity

Lower and upper quantiles respond differently to $X$.
:::
:::

. . .

```{r fig-homhet}
#| fig-width: 8.5
#| fig-height: 3
#| echo: false

n <- 300
x <- runif(n, 1, 10)

df_both <- bind_rows(
  tibble(x = x, y = 2 + 0.8*x + rnorm(n, 0, 2),
         scenario = "Homogeneous effect\n(parallel QR lines)"),
  tibble(x = x, y = 2 + 0.8*x + rnorm(n, 0, 0.5*x),
         scenario = "Heterogeneous effect\n(fanning QR lines)")
) |>
  mutate(scenario = factor(scenario, levels = c(
    "Homogeneous effect\n(parallel QR lines)",
    "Heterogeneous effect\n(fanning QR lines)"
  )))

qr_df <- map_dfr(levels(df_both$scenario), \(sc) {
  d <- filter(df_both, scenario == sc)
  map_dfr(taus, \(t) {
    fit <- rq(y ~ x, data = d, tau = t)
    tibble(scenario = factor(sc, levels = levels(df_both$scenario)),
           tau = factor(paste0("Q", t*100), levels = paste0("Q", taus*100)),
           intercept = coef(fit)[1], slope = coef(fit)[2])
  })
})

ggplot(df_both, aes(x, y)) +
  geom_point(alpha = 0.20, colour = "grey50") +
  geom_abline(data = qr_df, aes(intercept = intercept, slope = slope, colour = tau),
              linewidth = 0.9) +
  scale_colour_brewer(palette = "RdYlBu", name = "Quantile", direction = -1) +
  facet_wrap(~scenario) +
  labs(x = "Predictor", y = "Outcome")
```

# How Is It Estimated? {background-color="#2C3E50"}

## Three loss functions: three estimators

| Minimise | Estimator | Estimates |
|---|---|---|
| $\sum (y_i - \mu)^2$ | OLS | Conditional <span style="color:red;">mean</span> |
| $\sum \|y_i - \xi\|$ | Least absolute deviations | Conditional <span style="color:red;">median</span> |
| $\sum \rho_\tau(y_i - \xi)$ | Quantile regression | Conditional <span style="color:red;">$\tau$-th quantile</span> |

. . .

The key: **the criterion we minimise determines which part of the distribution we estimate.**

## Why does absolute loss give the median?

```{r fig-lad}
#| fig-height: 3.5
#| fig-width: 6
#| echo: false

households    <- c(100, 150, 200, 250, 800)
candidates    <- seq(50, 850, by = 1)
total_abs_dev <- sapply(candidates, \(m) sum(abs(households - m)))

tibble(candidate = candidates, total_abs_dev = total_abs_dev) |>
  ggplot(aes(x = candidate, y = total_abs_dev)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 200, colour = "#2ca25f", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 300, colour = "tomato",  linetype = "dashed", linewidth = 1) +
  annotate("text", x = 220, y = 1550, label = "Median = 200", colour = "#2ca25f", hjust = 0) +
  annotate("text", x = 320, y = 1450, label = "Mean = 300",   colour = "tomato",  hjust = 0) +
  labs(title = "Five households: 100, 150, 200, 250, 800",
       x = "Candidate value ξ", y = "Total absolute deviation")
```

The outlier at 800 pulls the **mean** away. The **median** is unaffected.

## The check function $\rho_\tau(u)$

$$\rho_\tau(u) = u \cdot \bigl(\tau - \mathbf{1}(u < 0)\bigr)$$

. . .

The term $\mathbf{1}(u < 0)$ is an **indicator function**: it equals <span style="color:red;">1</span> when the observation is <span style="color:red;">*below*</span> the fitted line, <span style="color:#1f77b4;">0</span> when it is <span style="color:#1f77b4;">*above*</span>.

. . .

::: {.columns}
::: {.column width="50%"}
**When $u > 0$** (above line):

penalty $= u \times \tau$
:::
::: {.column width="50%"}
**When $u < 0$** (below line):

penalty $= |u| \times (1 - \tau)$
:::
:::

. . .

Changing $\tau$ changes which side is penalised more heavily $\rightarrow$ the line moves to a different part of the distribution.

## The check function - visually

```{r fig-checkfn}
#| fig-height: 3.8
#| fig-width: 9
#| echo: false

u_vals     <- seq(-3, 3, length.out = 500)
tau_vals   <- c(0.10, 0.50, 0.90)
tau_labels <- c("τ = 0.10\n(10th percentile)", "τ = 0.50\n(Median)", "τ = 0.90\n(90th percentile)")

loss_df <- map_dfr(seq_along(tau_vals), \(i)
  tibble(u    = u_vals,
         loss = check_fn(u_vals, tau_vals[i]),
         tau  = factor(tau_labels[i], levels = tau_labels)))

loss_df |>
  ggplot(aes(x = u, y = loss)) +
  geom_line(colour = "steelblue", linewidth = 1.3) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50") +
  facet_wrap(~tau) +
  labs(x = "Residual  u = y − ŷ", y = "Loss  ρτ(u)",
       subtitle = "Steeper arm = heavier penalty on that side")
```

## The indicator function - concrete numbers

Seven observations, $\tau = 0.25$. **Which $\xi$ minimises the sum?**

$$\sum \rho_\tau(y_i - \xi)$$  

Quantile regression | Conditional <span style="color:red;">$\tau$-th quantile</span> 

## **Guess** $\xi = 0.674$ $\rightarrow$ Sum = 3.84 ❌

::: {.columns}
::: {.column width="97%"}

```{r tbl-bad}
#| echo: false

y7 <- c(1.622,-0.161,-0.461,-0.612,0.674,0.711,-0.865)
tibble(
  yi = round(y7, 3),
  `u` = round(y7 - 0.674, 3),
  `1(u<0)` = as.integer((y7 - 0.674) < 0),
  `⍴` = round(check_fn(y7 - 0.674, 0.25), 3)
) |> knitr::kable()
```
:::
:::

## **Optimal** $\xi = -0.612$ $\rightarrow$ Sum = 1.55 \ding{51} \checkmark

::: {.columns}
::: {.column width="97%"}

```{r tbl-opt}
#| echo: false

tibble(
  yi = round(y7, 3),
  `u` = round(y7 - (-0.612), 3),
  `1(u<0)` = as.integer((y7 - (-0.612)) < 0),
  `⍴` = round(check_fn(y7 - (-0.612), 0.25), 3)
) |> knitr::kable()
```
:::
:::

. . .

At the optimal $\xi$, exactly **1 of 7 observations** lies below, approximately 14%, close to the 25th percentile of this small sample.

## 🗂️ Hands-On Activity

::: {.callout-note}
Open the spreadsheet **`QR_CheckFunction.xlsx`**.

- Change $\xi$ manually → watch the total sum respond
- Change $\tau$ → see how the penalty weights shift
- Find the minimum by hand

When you have found it, `rq()` will do the same thing in one line.
:::

# Visual Intuition {background-color="#2C3E50"}

## The Engel data

Food expenditure vs household income, Belgium, 1857 ($n = 235$)

```{r load-engel}
#| echo: false
#| fig-height: 4

data(engel, package = "quantreg")

engel |>
  ggplot(aes(x = income, y = foodexp)) +
  geom_point(alpha = 0.4, colour = "grey40") +
  labs(x = "Annual income (Belgian francs)",
       y = "Annual food expenditure (Belgian francs)",
       subtitle = "Notice: the spread of food expenditure grows with income (heteroscedasticity)")
```

## OLS: one line through the data

```{r fig-ols}
#| echo: false
#| fig-height: 4

ols_engel <- lm(foodexp ~ income, data = engel)

engel |>
  ggplot(aes(x = income, y = foodexp)) +
  geom_point(alpha = 0.35, colour = "grey40") +
  geom_abline(intercept = coef(ols_engel)[1], slope = coef(ols_engel)[2],
              linewidth = 1.2) +
  labs(x = "Annual income", y = "Annual food expenditure",
       subtitle = "OLS estimates the conditional mean, but misses the spread")
```

## Quantile regression: a family of lines

```{r fig-fan}
#| echo: false
#| fig-height: 4.2

qr_lines <- map_dfr(taus, \(t) {
  fit <- rq(foodexp ~ income, data = engel, tau = t)
  tibble(tau = factor(paste0("Q", t*100), levels = paste0("Q", taus*100)),
         intercept = coef(fit)[1], slope = coef(fit)[2])
})

engel |>
  ggplot(aes(x = income, y = foodexp)) +
  geom_point(alpha = 0.30, colour = "grey50") +
  geom_abline(data = qr_lines,
              aes(intercept = intercept, slope = slope, colour = tau),
              linewidth = 1) +
  scale_colour_brewer(palette = "RdYlBu", name = "Quantile", direction = -1) +
  labs(x = "Annual income", y = "Annual food expenditure",
       subtitle = "The fan reveals heteroscedasticity that OLS averages over")
```

The lines spread at higher income levels, wealthier households differ more from each other.

## Can quantile lines cross?

**Theoretically:** No, $Q_\tau(Y \mid x)$ is non-decreasing in $\tau$ for fixed $x$.

**In practice:** Estimated lines *can* cross, especially with:

- Small samples
- Extreme quantiles ($\tau < 0.05$ or $> 0.95$)
- Flexible nonlinear models

This is undesirable because true quantiles should be non-decreasing in \tau.

## Estimated Quantiles Change in Steps

In finite samples, estimated quantile functions are step functions rather than perfectly smooth curves.

```{r fig-stepfn}
#| echo: false
#| fig-height: 4.5

df7    <- tibble(x = c(1,2,3,5,6,8,9), y = c(2,1,5,2,8,6,9))
xbar7  <- mean(df7$x)
tau_fine  <- seq(0.01, 0.99, by = 0.01)
q_at_xbar <- map_dbl(tau_fine, \(t)
  suppressWarnings(predict(rq(y ~ x, data = df7, tau = t),
                           newdata = data.frame(x = xbar7))))

tibble(tau = tau_fine, q = q_at_xbar) |>
  ggplot(aes(tau, q)) +
  geom_step(colour = "steelblue", linewidth = 1.2) +
  labs(title = "Q̂(τ | x̄) — always a step function",
       x = "τ", y = "Q̂(τ | x̄)",
       subtitle = "Non-decreasing by construction")
```
In large survey datasets, especially for central quantiles (e.g., \tau=0.25, 0.5, 0.75), quantile crossing is uncommon in practice because the estimates are relatively stable.

# `rq()` in R {background-color="#2C3E50"}

## The `rq()` function

Syntax mirrors `lm()`:

```{r rq-demo}
#| eval: true
#| code-fold: false

# Median regression (τ = 0.5)
fit_median <- rq(foodexp ~ income, data = engel, tau = 0.5)
summary(fit_median)
```

## Multiple quantiles at once

```{r rq-multi}
#| code-fold: false

fit_multi <- rq(
  foodexp ~ income,
  data = engel,
  tau  = c(0.10, 0.25, 0.50, 0.75, 0.90)
)
```

. . .

Returns an `rqs` object → one model per $\tau$.

. . .

::: callout-note
Default standard errors use **rank inversion**. For applied work, bootstrapped SEs (`se = "boot"`) are often preferred. This is covered in Section 04.
:::

# Interpreting Coefficients {background-color="#2C3E50"}

## The interpretation template

For $Q_\tau(\text{foodexp} \mid \text{income}) = \beta_0(\tau) + \beta_1(\tau) \times \text{income}$:

> *"At the $\tau$-th quantile of food expenditure, a one-unit increase in income is associated with $\hat{\beta}_1(\tau)$ francs, holding other variables constant."*

. . .

The phrase **"at the $\tau$-th quantile"** is not optional, it is what distinguishes QR from OLS.

## Slope across quantiles - Engel data

```{r coef-table}
#| echo: false

map_dfr(taus, \(t) {
  fit <- rq(foodexp ~ income, data = engel, tau = t)
  tibble(`τ` = t,
         Quantile = paste0(t*100, "th percentile"),
         `Slope β̂₁(τ)` = round(coef(fit)["income"], 4))
}) |> knitr::kable()
```

. . .

The slope increases across quantiles $\Rightarrow$ income matters **more** for high-spending households.

## Coefficient plot

```{r fig-coefplot}
#| echo: false
#| fig-height: 3.8

ols_slope <- coef(lm(foodexp ~ income, data = engel))["income"]

coef_plot_df <- map_dfr(taus, \(t) {
  fit <- rq(foodexp ~ income, data = engel, tau = t)
  s   <- summary(fit, se = "boot")
  cm  <- s$coefficients
  tibble(tau   = t,
         est   = cm["income", "Value"],
         lower = cm["income", "Value"] - 1.96 * cm["income", "Std. Error"],
         upper = cm["income", "Value"] + 1.96 * cm["income", "Std. Error"])
})

ggplot(coef_plot_df, aes(x = tau, y = est)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.20, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2.5) +
  geom_hline(yintercept = ols_slope, linetype = "dashed",
             colour = "tomato", linewidth = 0.9) +
  annotate("text", x = 0.12, y = ols_slope + 0.003,
           label = "OLS", colour = "tomato", size = 3.5, hjust = 0) +
  scale_x_continuous(breaks = taus, labels = paste0(taus*100, "th")) +
  labs(x = "Quantile (τ)", y = "Slope coefficient (income)",
       subtitle = "Shaded band = 95% bootstrap CI · Dashed line = OLS estimate")
```

# QR and OLS {background-color="#2C3E50"}

## When do they agree?

::: incremental
- **They agree** when the conditional distribution is symmetric and homoscedastic → OLS gives the full picture
- **They diverge** when:
  1. **Skewed outcomes**: wages, income, house prices. OLS is influenced by the upper tail; the median is not.
  2. **Heteroscedasticity**: the Engel data. Slopes differ systematically across quantiles.
  3. **Tail effects**: when the research question is specifically about high or low earners, extreme health outcomes, etc.
:::

## Take-home message

::: {.callout-important}
## Quantile regression complements OLS, it does not replace it.

If coefficients are **stable** across quantiles → OLS is telling you the full story.

If coefficients **vary** across quantiles → OLS is concealing important heterogeneity.

The coefficient plot across $\tau$ is your diagnostic.
:::

# What Comes Next {background-color="#2C3E50"}

## This afternoon: hands-on with CPS1988

We move to US wage data ($n = 28{,}155$, Current Population Survey 1988).

**Questions we will answer:**

::: incremental
- Does the return to an additional year of education differ for low and high earners?
- Is the ethnic wage gap uniform across the earnings distribution?
- How do we construct and interpret confidence intervals for QR estimates?
:::

# **Break now**: we will pick up with Session 04 (Inference) after the break. {background-color="#2C3E50"}

## 

::: {style="text-align: center; padding-top: 80px; font-size: 1.1em; font-style: italic; color: #555;"}
*"It is difficult to understand why statisticians limit their inquiries to Averages, and do not seem to revel in more comprehensive views."*

- Francis Galton, *Natural Inheritance*, 1890
:::
