---
title: "Case Study: Engel Curves and Quantile Regression"
subtitle: "Workshop Exercise"
author: "Tatjana Kecojević"
date: "24 June 2026"

format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show/Hide Code"
    smooth-scroll: true
    page-layout: article
 
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    documentclass: article
    geometry:
      - margin=1in

execute:
  echo: true
  warning: false
  message: false
---

# **Learning Objectives**

By the end of this exercise you will be able to:

- Fit quantile regression models using the Engel dataset
- Compare OLS and quantile regression estimates
- Examine how the relationship between income and food expenditure varies across quantiles
- Construct confidence intervals using `se = "nid"`
- Interpret a coefficient process plot
- Write up quantile regression findings in plain language

# **Introduction**

A classic question in economics is:

> How does household food expenditure change as income increases?

The Engel dataset is one of the most famous examples used in the quantile regression literature and is included in the `quantreg` package.

Unlike the CPS1988 wage example, this dataset contains only two variables:

- Household income
- Household food expenditure

This allows us to focus entirely on understanding quantile regression.

```{r ex0-setup}
#| include: false

library(tidyverse)
library(quantreg)
library(broom)

theme_set(theme_minimal())

data(engel, package = "quantreg")
```

## Exercise 1: Exploring the Data

Inspect the dataset

```{r ex0-inspect}
glimpse(engel)
summary(engel)
```

**Questions**

1. How many households are included?
2. Which variable is the outcome?
3. Which variable is the predictor?

---

## Visualising the Variables

```{r ex1-income}
ggplot(engel, aes(x = income)) +
  geom_histogram(bins = 30,
                 fill = "steelblue",
                 alpha = 0.7) +
  labs(
    title = "Household Income",
    x = "Income",
    y = "Count"
  )
```


```{r ex1-foodexp}
ggplot(engel, aes(x = foodexp)) +
  geom_histogram(bins = 30,
                 fill = "steelblue",
                 alpha = 0.7) +
  labs(
    title = "Household Food Expenditure",
    x = "Food expenditure",
    y = "Count"
  )
```

**Questions**

1. Is income skewed?
2. Is food expenditure skewed?
3. Why might averages be insufficient to describe these distributions?

## Exercise 2: Income and Food Expenditure

```{r ex2-scatter}
ggplot(
  engel,
  aes(
    x = income,
    y = foodexp
  )
) +
  geom_point(alpha = 0.5,
             colour = "steelblue") +
  labs(
    title = "Food Expenditure and Household Income",
    x = "Income",
    y = "Food expenditure"
  )
```

**Questions**

1. Is the relationship positive?
2. Does variability appear constant across the income range?
3. Why might quantile regression be useful here?

## Exercise 3: OLS Baseline

```{r ex3-ols}
ols_fit <- lm(
  foodexp ~ income,
  data = engel
)

summary(ols_fit)
```

**Questions**

1. What is the coefficient on income?
2. Interpret the coefficient.
3. What does OLS tell us about the relationship between income and food expenditure?

## Exercise 4: Median Regression

```{r ex4-median}
fit_median <- rq(
  foodexp ~ income,
  tau = 0.50,
  data = engel
)

summary(
  fit_median,
  se = "nid"
)

# Alternative:
# summary(
#   fit_median,
#   se = "boot",
#   R = 1000
# )
```

**Questions**

1. How does the median regression estimate compare with OLS?
2. Why might the two estimates differ?
3. What does median regression estimate that OLS does not?

## Exercise 5: Multiple Quantiles

```{r ex5-multi}
taus <- c(
  0.10,
  0.25,
  0.50,
  0.75,
  0.90
)

fit_multi <- rq(
  foodexp ~ income,
  tau = taus,
  data = engel
)

coef(fit_multi)
```

**Questions**

1. Is the income coefficient constant across quantiles?
2. Which quantile has the largest coefficient?
3. What does this imply about spending behaviour?

### Formal Test of Equality Across Quantiles

```{r ex5-anova}

anova(fit_multi)

```

**Questions**

4. What is the null hypothesis being tested?
5. What does the p-value suggest?
6. Is there evidence that the income effect differs across quantiles?

## Exercise 6: Confidence Intervals

```{r ex6-ci}
income_df <- map_dfr(taus, \(t){

  fit <- rq(
    foodexp ~ income,
    tau = t,
    data = engel
  )

  s <- summary(
    fit,
    se = "nid"
  )

  # Alternatives:
  # s <- summary(fit, se = "boot", R = 50)    # quick experimentation
  # s <- summary(fit, se = "boot", R = 1000)  # more stable bootstrap inference
  
  cm <- s$coefficients

  tibble(
    tau   = t,
    est   = cm["income","Value"],
    lower = cm["income","Value"] -
      1.96 * cm["income","Std. Error"],
    upper = cm["income","Value"] +
      1.96 * cm["income","Std. Error"]
  )
})

income_df
```

**Questions**

1. Are all estimated income effects positive?
2. How do the confidence intervals change across quantiles?
3. Do the intervals suggest that the effect of income is constant?
4. Why are confidence intervals useful when interpreting quantile regression results?

## Exercise 7: Coefficient Plot

```{r ex7-plot}
ols_income <- coef(ols_fit)["income"]

ggplot(
  income_df,
  aes(
    x = tau,
    y = est
  )
) +
  geom_ribbon(
    aes(
      ymin = lower,
      ymax = upper
    ),
    alpha = 0.20,
    fill = "steelblue"
  ) +
  geom_line(
    colour = "steelblue",
    linewidth = 1
  ) +
  geom_point(
    colour = "steelblue",
    size = 2.5
  ) +
  geom_hline(
    yintercept = ols_income,
    linetype = "dashed",
    colour = "tomato"
  ) +
  labs(
    title = "Income Effect Across Quantiles",
    subtitle = "Blue = Quantile Regression · Red dashed = OLS",
    x = "Quantile (τ)",
    y = "Coefficient on income"
  )
```

**Questions**

1. What pattern do you observe?
2. Does the effect of income appear constant?
3. Does OLS adequately summarise the relationship?

## Exercise 8: Interpretation

Complete the following paragraph.

At the 10th quantile, an additional unit of income is associated with approximately _____ additional units of food expenditure. At the 90th quantile, the corresponding estimate is _____. This suggests that ________________________________________. The OLS estimate of _____ provides _____________________________________________.

---

**Reflection**

1. What did quantile regression reveal that OLS could not?
2. Which result was most surprising?
3. In what other applications might quantile regression be useful?